---
title: "Demo R Notebook"
output: html_notebook
---

## Background
Here I've downloaded methylation profiling data (Illumina Infinium) from GEO, dataset can be found [here](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63409). This study took bone marrow samples from 5 healthy patients, and blood samples from 15 patients with blood cancer.

For the healthy, the authors sorted out hematopoeietic stem cells (HSCs) and progenitor cells (LMPP, MPP, etc) based on external markers with FACS.
For the leukemia patients, the authors sorted out putative leukemia stem cells (LSCs, CD34+CD38-), and blasts (CD34-), and other (CD34+CD38+) based on CD34 and CD38 expression with FACS.

My project found that a single genetic mutation in Tet2 dramatically changes leukemia blasts to act more like leukemia stem cells. I was curious if my cells' methylation profiles were more similar to explicitly profiled LSCs following the Tet2 mutation.

Prior to what you see here, I used the ChAMP library to normalize the methylation data (quantile-based), and made a table of metadata.


## Attach required packages and get data (local)
```{r, messages=F, set.seed(5)}
knitr::opts_chunk$set(cache = T)

# attach required packages
library(ggplot2)
library(matrixStats)
library(FactoMineR)
library(MASS)
library(RColorBrewer)
library(dendextend)
library(reshape2)

# read in normed methylation data from Jung et al
bvals = readRDS("../data/Jung_champnorm.rds")
nsamples = ncol(bvals)

# read in phenotype metadata
pheno_Jung = readRDS(paste0("../data/Jung_pheno.rds"))
pheno = data.frame(dataset = "Jung",
                   cell.type = as.character(pheno_Jung$Subject.ID),
                   sample = pheno_Jung$Sample_Name,
                   patient = pheno_Jung$Human)
pheno$sample.type = ifelse(pheno$cell.type %in% c("CD34-", "CD34+38+", "CD34+38-"), "Cancer", "Normal")

# set cell type factor levels for proper ordering
pheno$cell.type = factor(pheno$cell.type, 
                          levels=c("HSC", "MPP", "LMPP", "CMP", "GMP", "MEP", 
                                   "CD34+38-", "CD34+38+", "CD34-"))

# set up palette for consistent colors across figures
pal = c("tomato3", "sienna2", "darkgoldenrod1", "darkolivegreen2", "springgreen3", "mediumseagreen", 
                 "darkorchid4", "darkorchid1", "deeppink1")
names(pal) = levels(pheno$cell.type)
patientpal = colorRampPalette(brewer.pal(9, "Set1"))(20)
names(patientpal) = levels(pheno$patient)

```



## PCA of all samples 
For a nice overview of all samples, I start with PCA and color points based on cell type. Qualitatively, the normal cells (red-green) are already separate from the leukemic cells (pink/purple). However, the cell types are intermixed in this projection.
```{r}
# # set seed
# set.seed(5)

# sample 10,000 CpGs to plot
bvals_small = bvals[sample(1:nrow(bvals), 10000), ]

# find top 3 PCs with PCA of smaller dataset (~8s)
num_dimensions_PCA = 3
bVal_PCA = PCA(t(bvals_small), ncp=num_dimensions_PCA, scale.unit=T, graph=F)

# get transformed coords, and add metadata for simpler ggplot visualization
plot_PCA = as.data.frame(bVal_PCA$ind$coord)
names(plot_PCA) = paste0("PC", 1:num_dimensions_PCA)
plot_PCA$dataset = pheno$dataset
plot_PCA$cell.type = pheno$cell.type

# scatterplot of PC1 v PC2 colored by cell type
ggplot(plot_PCA, aes(x=PC1, y=PC2, color=cell.type)) +
  geom_point() + theme_classic() +
  scale_color_manual(values=pal) + labs(color="Cell type")

```


## Heatmap of leukemia cells
Ultimately, I want to know if my cells with a Tet2 mutation are more LSC-like or not. So, from this dataset, I want to know methylation features make a cell an LSC.
However, hierarchical clustering of the leukemia cells in this dataset show that differences between patients are larger than differences between cell type, making it difficult to initially understand what makes LSCs different.
```{r}
# sample 1000 CpGs to plot in heatmap
bvals_heatmap = bvals_small[sample(1:nrow(bvals_small), 1000), ]

# how do patient cancer samples cluster
# AKA is it easy to find out what separates cancer stem cells from proliferative cancer cells?
leuk_samples = pheno$sample.type == "Cancer"
leuk_cluster_data = t(bvals_heatmap[,leuk_samples])
leuk_cluster_pheno = pheno[leuk_samples,]
leuk_clusters = as.dendrogram(stats::hclust(dist(leuk_cluster_data, method="euclidean"), method="ward.D2"))

# change colors
morecols = colorRampPalette(c("dodgerblue4", "gray0", "firebrick4"))

# show heatmap
gplots::heatmap.2(leuk_cluster_data, 
                  srtCol = 20,
                  dendrogram = "both",
                  Rowv = leuk_clusters,
                  Colv = T,
                  labCol = NA,
                  ylab = "patients",
                  xlab = "CpGs",
                  labRow = NA,
                  trace = "none",          
                  margins = c(2,2),    
                  key.xlab = "beta value",
                  key.title = "Methylation",
                  denscol = "grey",
                  density.info = "density",
                  RowSideColors = pal[leuk_cluster_pheno[,"cell.type"]], # to add nice colored strips
                  col = morecols(13)
                  )

```
**Key:**  
Dark purple: *HSCs*,  
Light purple: *CD34+CD38+*,  
Pink: *CD34- blasts*


### Heatmap of leukemia cells colored by patient
```{r}
gplots::heatmap.2(leuk_cluster_data, 
                  srtCol = 20,
                  dendrogram = "both",
                  Rowv = leuk_clusters,
                  Colv = T,
                  labCol = NA,
                  ylab = "patients",
                  xlab = "CpGs",
                  labRow = NA,
                  trace = "none",          
                  margins = c(2,2),    
                  key.xlab = "beta value",
                  key.title = "Methylation",
                  denscol = "grey",
                  density.info = "density",
                  RowSideColors = patientpal[leuk_cluster_pheno$patient], # to add nice colored strips
                  col = morecols(13)
                  )

```


### Calculate cluster purity for known cancer cell types
```{r}
cut_cluster = melt(cutree(leuk_clusters, k = 3), value.name="cluster")
cut_cluster$sample = rownames(cut_cluster)
pheno_cluster = merge(pheno, cut_cluster, by="sample")

confusion_mat = table(pheno_cluster[,c("cluster", "cell.type")])
initial_purity_cancer = sum(rowMaxs(confusion_mat))/nrow(pheno_cluster)
initial_purity_cancer

```


## Heatmap of normal bone marrow cells
In normal cells, too, cells from the same patient are qualitatively more similar to each other than cells of the same type across patients. Although, hierarchical clustering shows some samples are not predominantly grouped by patient.
```{r}
# How do normal cells cluster?
normal_samples = pheno$sample.type == "Normal"
norm_cluster_data = t(bvals_heatmap[,normal_samples])
norm_cluster_pheno = pheno[normal_samples,]
norm_clusters = as.dendrogram(stats::hclust(dist(norm_cluster_data, method="euclidean"), method="ward.D2"))

morecols = colorRampPalette(c("dodgerblue4", "gray0", "firebrick4"))

gplots::heatmap.2(norm_cluster_data, 
                  srtCol = 20,
                  dendrogram = "both",
                  Rowv = norm_clusters,
                  Colv = T,
                  labCol = NA,
                  ylab = "patients",
                  xlab = "CpGs",
                  labRow = NA,
                  trace = "none",          
                  margins = c(2,2),     
                  key.xlab = "beta value",
                  key.title = "Methylation",
                  denscol = "grey",
                  density.info = "density",
                  RowSideColors = pal[norm_cluster_pheno[,"cell.type"]], # to add nice colored strips
                  col = morecols(13)
                  )
```
**Key:**  
Red: *HSCs*,  
Orange: *MPP*,  
Yellow: *LMPP*,   
Light green: *CMP*,   
Green: *GMP*,  
Dark green: *MEP*

### Heatmap of normal cells colored by patient
```{r}
gplots::heatmap.2(norm_cluster_data, 
                  srtCol = 20,
                  dendrogram = "both",
                  Rowv = norm_clusters,
                  Colv = T,
                  labCol = NA,
                  ylab = "patients",
                  xlab = "CpGs",
                  labRow = NA,
                  trace = "none",          
                  margins = c(2,2),    
                  key.xlab = "beta value",
                  key.title = "Methylation",
                  denscol = "grey",
                  density.info = "density",
                  RowSideColors = patientpal[norm_cluster_pheno$patient], # to add nice colored strips
                  col = morecols(13)
                  )

```


### Calculate cluster purity for known normal cell types
```{r}
cut_cluster = melt(cutree(norm_clusters, k = 6), value.name="cluster")
cut_cluster$sample = rownames(cut_cluster)
pheno_cluster = merge(pheno, cut_cluster, by="sample")

confusion_mat = table(pheno_cluster[,c("cluster", "cell.type")])
initial_purity_normal = sum(rowMaxs(confusion_mat))/nrow(pheno_cluster)
initial_purity_normal

```


## Use LDA to separate HSCs, LSCs, and blasts
```{r}
# Use LDA to separate HSCs, LSCs, and differentiated tumor cells
bvals_LDA = t(scale(bvals))
train_grps = pheno$cell.type %in% c("HSC", "CD34-", "CD34+38-")
bvals_train = bvals_LDA[train_grps,]

# find top 10,000 most variable CpGs, subset betas
bvals_summary = data.frame(mean=colMeans(bvals_train),
                           stdev=colSds(as.matrix(bvals_train)))
bvals_summary = bvals_summary[order(bvals_summary$stdev, decreasing=T),]
bvals_train = bvals_train[,rownames(bvals_summary[1:10000,])]

# add cell type as a metadata column, set factor order
bvals_train = data.frame(bvals_train, cell.type = pheno$cell.type[train_grps])
bvals_train$cell.type = factor(as.character(bvals_train$cell.type), 
                               levels = c("HSC", "CD34-", "CD34+38-"))

# LDA (~30s)
# ptm = proc.time()
dfLDA = lda(formula = cell.type ~ .,
            data=bvals_train)
# proc.time() - ptm

# calculate amt of between group variance that is explained by each LD
prop = dfLDA$svd^2/sum(dfLDA$svd^2)
prop 

# project other samples into LDA transformed space (~30s)
plda = predict(object=dfLDA,
               newdata=as.data.frame(bvals_LDA))

# scatterplot of transformed data
ggplot(data=as.data.frame(plda$x), aes(x=LD1, y=LD2, color=pheno$cell.type)) + 
  geom_point() +
  scale_color_manual(values=pal) + 
  theme_classic() + 
  labs(color="Cell type")


```


## Cluster cancer samples in the LDA-transformed space
```{r, messages=F}
# how do patient cancer samples cluster in the new space?
leuk_samples = pheno$sample.type == "Cancer"
leuk_cluster_data = as.matrix(plda$x)[leuk_samples,]
leuk_cluster_pheno = pheno[leuk_samples,]
leuk_clusters = as.dendrogram(stats::hclust(dist(leuk_cluster_data, method="euclidean"), method="ward.D2"))

# change colors
morecols = colorRampPalette(c("dodgerblue4", "gray0", "firebrick4"))
nclust = 3
leuk_clusters = color_branches(leuk_clusters, k = nclust)

# show heatmap
gplots::heatmap.2(leuk_cluster_data, 
                  srtCol = 20,
                  dendrogram = "both",
                  Rowv = leuk_clusters,
                  Colv = T,
                  labCol = NA,
                  ylab = "patients",
                  xlab = "LDs",
                  labRow = NA,
                  trace = "none",          
                  margins = c(2,2),    
                  key.xlab = "beta value",
                  key.title = "Methylation",
                  denscol = "grey",
                  density.info = "density",
                  RowSideColors = pal[leuk_cluster_pheno[,"cell.type"]], # to add nice colored strips
                  col = morecols(13)
                  )

```
**Key:**  
Dark purple: *HSCs*,  
Light purple: *CD34+CD38+*,  
Pink: *CD34- blasts*

### Calculate cluster purity
```{r}
cut_cluster = melt(cutree(leuk_clusters, k = nclust), value.name="cluster")
cut_cluster$sample = rownames(cut_cluster)
pheno_cluster = merge(pheno, cut_cluster, by="sample")

confusion_mat = table(pheno_cluster[,c("cluster", "cell.type")])
lda_purity_cancer = sum(rowMaxs(confusion_mat))/nrow(pheno_cluster)
lda_purity_cancer

```



```{r}
# How do normal cells cluster in the new space?
normal_samples = pheno$sample.type == "Normal"
norm_cluster_data = as.matrix(plda$x)[normal_samples,]
norm_cluster_pheno = pheno[normal_samples,]
norm_clusters = as.dendrogram(stats::hclust(dist(norm_cluster_data, method="euclidean"), method="ward.D2"))

# change colors
morecols = colorRampPalette(c("dodgerblue4", "gray0", "firebrick4"))
nclust = 6
norm_clusters = color_branches(norm_clusters, k = nclust)

# show heatmap
gplots::heatmap.2(norm_cluster_data, 
                  srtCol = 20,
                  dendrogram = "both",
                  Rowv = norm_clusters,
                  Colv = T,
                  labCol = NA,
                  ylab = "patients",
                  xlab = "LDs",
                  labRow = NA,
                  trace = "none",          
                  margins = c(2,2),    
                  key.xlab = "beta value",
                  key.title = "Methylation",
                  denscol = "grey",
                  density.info = "density",
                  RowSideColors = pal[norm_cluster_pheno[,"cell.type"]], # to add nice colored strips
                  col = morecols(13)
                  )

```
**Key:**  
Red: *HSCs*,  
Orange: *MPP*,  
Yellow: *LMPP*,   
Light green: *CMP*,   
Green: *GMP*,  
Dark green: *MEP*


### Calculate cluster purity for known normal cell types
```{r}
cut_cluster = melt(cutree(norm_clusters, k = nclust), value.name="cluster")
cut_cluster$sample = rownames(cut_cluster)
pheno_cluster = merge(pheno, cut_cluster, by="sample")

confusion_mat = table(pheno_cluster[,c("cluster", "cell.type")])
lda_purity_normal = sum(rowMaxs(confusion_mat))/nrow(pheno_cluster)
lda_purity_normal

```


## Show cluster purities
```{r}
paste("Initial normal cluster purity:", round(initial_purity_normal, digits=4))
paste("LDA normal cluster purity:", round(lda_purity_normal, digits=4))

paste("Initial cancer cluster purity:", round(initial_purity_cancer, digits=4))
paste("LDA cancer cluster purity:", round(lda_purity_cancer, digits=4))

```


